A generalized definition of multidimensional item response theory parameters

### 2024-09-03 output/paper_draft/paper_M2PLM_parameters.Rmd

We must stop here and make a few remarks. First, as we are not constraining
ourselves to cognitive traits,
we speak of _positive_ instead of _correct response_. This may imply giving
a correct response indeed, if we are talking about a cognitive test.
But it may mean endorsing an item (or a certain response option)
in a non-cognitive domain.
Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
non-negative real numbers in Reckase
[-@reckase_difficulty_1985, a decrease in the probability of a correct response
with an increase in ability makes little sense].
In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
of the $`r DIM_INDEX`$-th component $`r DISCR_PARAM`$ can be negative if
the item is *inverse* for (i.e., taps the negative pole of)
latent dimension $`r DIM_INDEX`$ (e.g., a negatively-keyed item).
Finally, as responses are not necessarily interpreted as correct or incorrect,
parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_
instead of the item difficulty.

Following @reckase_difficulty_1985’s procedure,
we must find in the first place
the point of maximum slope of $`r IRF_ABBR`$ in $`r LATENT_SPACE`$.
Because the slope is dependent on the direction considered,
we need to reparameterize Equation \@ref(eq:M2PL-formulation)
in polar coordinates, as in @reckase_difficulty_1985.
However, Reckase's Equation 3 implicitly assumes orthogonality.
To contemplate the general case,
we need to introduce a few linear algebra results before proceeding further.

@@

Thus, inner product $`r LS_BASIS_INNER_PROD`$ has $`r INNER_PROD_MAT_EQ`$ as
Gram matrix, i.e., $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$.

### Test space definition

Let $`r TEST_SPACE`$ be the vector space made up by
the set of $`r DISCR_VECTOR`$ vectors,
with the operations sum and product by a scalar as in $`r REAL_N_SPACE`$.
We want @reckase_difficulty_1985's definition of
the multidimensional parameters in rectangular coordinates to be
a particular case of our more general definition;
we can thus consider an orthonormal basis $`r TEST_SPACE_STD_BASIS`$
in $`r TEST_SPACE`$ such that $`r DISCR_STD_COORDS_ALT_EQ`$
corresponds in the model to $`r TRAIT_ORTH_COORDS`$.
Since the response probability to the `r MODEL_ACRONYM`,
given by Equation \@ref(eq:M2PL-formulation), must be invariant with respect to
the change of basis expressed in Equation \@ref(eq:change-basis),
$`r INVARIANCE_EQ`$.
Therefore, we find a basis $`r TEST_SPACE_BASIS_EQ`$ in $`r TEST_SPACE`$
defined by

\begin{equation}
  `r DISCR_STD_COORDS_EQ`,                               (\#eq:discr-vector-std)
\end{equation}

with $`r DISCR_COORDS_ALT_EQ`$, and
$`r TRANSFORM_MATRIX_TRANSP_INV`$ a change of basis matrix.
Note that $`r TRANSFORM_MATRIX_TRANSP_INV`$ determines $`r TEST_SPACE_BASIS`$.

To make @reckase_difficulty_1985's definition a particular case of ours
we can conveniently assume that the inner product in
$`r TEST_SPACE_STD_BASIS`$ is the dot product,
as we did with $`r LS_ORTH_BASIS`$.
As $`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.
The norm of $`r DISCR_VECTOR`$ thus fulfills

\begin{equation}
  `r DISCR_VECTOR_MODULE_EQ`.
\end{equation}

Therefore, an inner product $`r TEST_SPACE_INNER_PROD`$ with Gram matrix
$`r INNER_PROD_MATRIX_INV`$ can be defined, such that $`r INNER_PROD_DISCR_EQ`$
keeps the norm of $`r DISCR_VECTOR`$ invariant.
Following @zhang_theoretical_1999, we shall refer to space $`r TEST_SPACE`$ as
the *test space*, as it is the space where
the items (i.e., the test elements) are represented.

### Direction cosines in the latent space

@@

i. $`r TRAIT_VEC_ORTH_POLAR_EQ`$

i. $`r TRAIT_VECTOR_POLAR_EQ`$

i. $`r TRAIT_VEC_ALT_POLAR_EQ`$

Given i,

@@

Using ii and iii we get that

@@

## Relationship between test space and latent space

Before presenting the results about the parameters,
we must make a few remarks regarding the vector space $`r TEST_SPACE`$.
As shown before, this space is an entity by itself,
distinct from the latent space $`r LATENT_SPACE`$.
Despite this, the dot product in the exponent of Equation
\@ref(eq:M2PL-formulation) induces a bijective relationship between
their two sets of respective bases.
Explicitly, if $`r TRAIT_VECTOR`$ is represented in basis $`r LS_BASIS`$
with inner product Gram matrix $`r INNER_PROD_MATRIX`$,
$`r DISCR_VECTOR`$ is represented in basis $`r TEST_SPACE_BASIS`$
with inner product Gram matrix $`r INNER_PROD_MATRIX_INV`$,
for any pair of bases $`r LS_BASIS`$ and $`r TEST_SPACE_BASIS`$.
Only when the two bases are orthonormal,
the two spaces share a common inner product (i.e., the dot product),
and thus can be mistaken.
However, it is worth noting that, although their representation
can be superimposed in this case,
they are still two different spaces. <!-- # TODO: To discussion? -->

We must make a few observations here.
First, the transformation to the orthonormal bases of the two spaces allows
mapping the $`r MODEL_ACRONYM`$ parameters indistinctly in the two spaces,
by projecting one space onto the other.
This allows representing the coordinates of the discrimination vectors
in the latent space and, conversely,
the coordinates of the latent trait vectors in the test space.
As we can see, the invariance of the model holds,
no matter in which space we represent the coordinates.
<!-- # TODO: Incluir la condición de invarianza de las coordenadas en los dos espacios (Foto pizarra)? -->
Second, and most importantly, if the test space is
where the item parameters naturally lie, they must be represented in this space,
and not in the latent space.
Therefore, we may find the direction angles of $`r DISCR_VECTOR`$
in the test space:
As $`r LS_ORTH_BASIS`$ and $`r TEST_SPACE_STD_BASIS`$ are both
orthonormal in $`r REAL_N_SPACE`$, without loss of generality
we can consider that $`r BASES_EQ`$ (save for a rotation).
We can then map the latent space directly onto the test space and
apply Proposition 2, so if $`r ANGLE_VEC_TS_ITEM`$ is
the vector of direction angles of $`r DISCR_VECTOR`$,
in the test space $`r TEST_SPACE`$ we get

@@

with $`r DIAG_MATRIX_INNER_PROD_INV`$ a diagonal matrix where
$`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.
As we will see later, this representation is very convenient, as
the direction of the item vectors with respect to the coordinate axes
has a direct, meaningful interpretation.

@@

The second property derives straightforwardly from
Equation \@ref(eq:distance-param),
and the third one is implicit in the definition of the $`r MDISC_SYM`$.
However, when an item measures only dimension $`r DIM_INDEX`$,
simple arithmetic can show that $`r MDISC_UNIDIM_GENERALIZED_EQ`$.
In the covariance-based case, this means that $`r MDISC_COV_BASED_UNIDIM_EQ`$
and, as $`r COV_SYM_DIAG`$ is generally different from 1,
$`r MDISC_COV_PARAM`$ does not fulfill the first property
(we shall refer to this property as *scale invariance*, from now on).
Nevertheless, it can be interpreted as a *scaled version* of the parameter,
with the standard deviation of the corresponding dimension as scaling factor.
This transformation may be made in the unidimensional case as well,
thus finding a *standard-metric* discrimination parameter
(i.e., referred to a unitary variance space).

### Correlation-based version

Despite $`r MDISC_COV_PARAM`$ not being scale-invariant,
we have argued that it is still a valid generalization of the IRT parameters.
However, it may be convenient to find yet another generalization,
one that accounts for the latent space structure
while still fulfilling this property.
To do this, let's assume a matrix $`r TRANSFORM_MATRIX_AUX_EQ`$, such that
$`r TRANSFORM_MATRIX_AUX_RESULT`$.
Then, we have that

$$
  `r CORR_BASED_DERIVATION`.
$$

That is, defining $`r TRANSFORM_MATRIX`$ as
$`r TRANSFORM_MATRIX_CORR_BASED_DEF`$ can be interpreted as re-scaling
the transformed parameter back to its original metric.
In this case, we get that $`r INNER_PROD_CORR_COND`$.
Therefore, we can also define a *correlation-based* version of the
multidimensional parameters as

As we can see, this version also fulfills the scale invariance property,
as the $`r INNER_PROD_MAT_INV_DIAG_ELEMENT`$ elements are
the diagonal elements of $`r CORR_MATRIX`$, which are always equal to 1.
It is important to note though that this transformation does not have
an also desirable property,
which we have used to justify the covariance-based version:
The $`r TRAIT_ORTH_COORDS`$ coordinates resulting from transforming
$`r TRAIT_COORDS`$ to the orthonormal basis $`r LS_ORTH_BASIS`$
are not generally uncorrelated, as one would expect
(save some exceptions, e.g., when all the variances are equal).
Nevertheless, it is important to note how the $`r MDISC_CORR_PARAM`$ accounts
for the correlation structure of the latent space
while still being scale invariant.
This will be very useful for the purpose of representing the items graphically,
as we will see later.

## Vector representation

According to @ackerman_creating_1994, the most appropriate way of representing
multidimensional items is in vector form.
This allows representing several items altogether and analyzing them
in terms of their multidimensional parameters [@ackerman_graphical_1996].
An item vector will be applied at a direction and (signed) distance
from the origin of the coordinate system given by its $`r MIL_PARAM`$ parameter,
and will have a length equal to its $`r MDISC_SYM`$ parameter.
Its direction will also be given by the direction component of $`r MIL_PARAM`$,
so its orientation will always pass through the origin.
Based on these conditions, we need to compute
the origin and end coordinates of the vector,
denoted by $`r ORIGIN_ITEM_BASIS`$ and $`r END_ITEM_BASIS`$, respectively.

In an orthogonal basis it is easy to compute the coordinates,
as they are simply the orthogonal projections onto the corresponding axis.
In the oblique case it is not so simple though, but we can take advantage of the
orthornormalization introduced above.
In basis $`r TEST_SPACE_STD_BASIS`$, the origin and end coordinates are
$`r ORIGIN_ITEM_ORTH`$ and $`r END_ITEM_ORTH`$, respectively,
and they correspond to the coordinates in $`r TEST_SPACE_BASIS`$ transformed
according to Equation \@ref(eq:discr-vector-std),
i.e., $`r ORIGIN_TRANSF_EQ`$ and $`r END_TRANSF_EQ`$.
Given the mapping of $`r TEST_SPACE_STD_BASIS`$ onto $`r LS_ORTH_BASIS`$,
the direction cosines in $`r TEST_SPACE_STD_BASIS`$
are $`r DIR_COS_ITEM_VEC_STD`$.
Therefore,

\begin{align}
  `r ORIGIN_ORTH_EQ` \\
  `r END_ORTH_EQ`.
\end{align}

Applying Equation \@ref(eq:dir-cos-vec-std),
for the origin coordinates we have that

\begin{equation}
  `r ORIGIN_ORTH_DISCR_EQ`                              (\#eq:origin-orth-discr)
\end{equation}

which, operating, results in

\begin{align}
  `r ORIGIN_ITEM_COORDS_EQ`    \\
  `r ORIGIN_TRANSF_SQUARED_EQ` \\
  `r ORIGIN_INNERPROD_EQ`      \\
  `r ORIGIN_INNERPROD_DIAG_EQ`
\end{align}

and, finally,

\begin{equation}
  `r ORIGIN_INNERPROD_DIR_COS_EQ`.                   (\#eq:origin-coords-result)
\end{equation}

The end coordinates can be computed by a similar procedure, which results in

\begin{equation}
  `r END_ITEM_COORDS_EQ`.                               (\#eq:end-coords-result)
\end{equation}

Equations \@ref(eq:origin-coords-result) and \@ref(eq:end-coords-result)
express the items coordinates in terms of their multidimensional parameters.
However, it is worth noting that the latent space Gram matrix
also plays a role in computing these coordinates,
and will therefore have an effect on the representation.
Of course,
from Equations \@ref(eq:origin-orth-discr) and \@ref(eq:end-coords-result)
it is easy to express the coordinates in terms of the parameters in the original
model formulation, namely,

\begin{align}
  `r ORIGIN_ITEM_PARAMS_EQ`                  &(\#eq:origin-from-model-params) \\
  `r END_ITEM_PARAMS_EQ`.                    &(\#eq:end-from-model-params)
\end{align}

However, formulating them in terms of the multidimensional parameters allows us
studying the items in terms of their multidimensional properties.
In the following, we see how to geometrically interpret
these multidimensional parameters.

### Geometric properties of the items

We have seen how the two new versions of the parameters fulfill
the three conditions proposed by @reckase_discriminating_1991
for valid generalizations of the IRT parameters to the multidimensional case,
being the scale invariance of the $`r MDISC_COV_PARAM`$ the only exception.
From a geometrical perspective, the item $`r MDISC_SYM`$ is simply interpreted
as the length of its corresponding vector.
Regarding the $`r MIL_PARAM`$ parameter,
we will consider now two more properties, related to each of its two components:
the distance to the origin, and the measurement direction.

The (signed) distance to the origin,
given by the first component of $`r MIL_PARAM`$, has an interesting property:
By the definition of the inner product,
$`r MDISC_SYM`$ is strictly positive for non-null vectors.
Therefore, the sign of $`r DISTANCE_PARAM`$ is equal to
the sign of $`r INTERCEPT_PARAM`$.
The implication is that the item is displaced along
its measurement direction forwards or backwards with respect to the origin.
As the $`r MDISC_SYM`$ appears in the denominator of $`r DISTANCE_PARAM`$,
that displacement will be inversely proportional
to the discrimination of the item.
That is, the less discriminating the item is,
the further away it will be from the origin.

The measurement direction is given by the signs of $`r DIR_COS_ITEM_VEC`$ which,
also due to the strict positiveness of $`r MDISC_SYM`$,
are equal to the signs of $`r DISCR_VECTOR`$.
This property leads directly to generalizing
the parameters to the monotonically non-increasing case:
The measurement direction relative to dimension $`r DIM_INDEX`$
will be negative when $`r IRF_ABBR`$ is monotonically decreasing
with respect to variations along $`r TRAIT_COMPONENT`$.
When $`r IRF_ABBR`$ is constant
with respect to variations along $`r TRAIT_COMPONENT`$,
$`r SIGN_COS_VEC_ITEM_EQ`$,
which means that $`r MIL_PARAM`$ is orthogonal to the $`r DIM_INDEX`$-th axis.
However, note that this does not necessarily imply that
the vector is parallel to any other axis
(or strictly contained in the hyperplane formed by other axes, for that matter);
this will always be true for the $`r MIL_AG_PARAM`$ due to
the orthogonality assumption,
but it will depend on the correlation matrix $`r CORR_MATRIX`$
for the other two versions of the parameter.

@@

The geometric properties are easier to apprehend with
a visual representation of the items.
To do that, we need to choose a version of the parameters
that yields their most faithful representation.
We have already seen that
the agnostic version disregards the latent space structure.
Therefore, it will only be useful in the unlikely case that
we have no information of such structure.
When we have an estimate of the latent space covariances,
we may use either the covariance-based or the correlation-based version.
However, we have already seen how
the distance to the origin depends on the $`r MDISC_SYM`$.
Also, in the covariance-based,
the term $`r INNER_PROD_MAT_DIAG_INV_PROD`$ becomes $`r COV_BASED_SCALE_MAT_EQ`$
which, as we have argued before, implies a change of scale.
In the case of the correlation-based version however,
$`r INNER_PROD_MAT_DIAG_INV_PROD`$ is simply $`r CORR_MATRIX_INV`$,
being the coordinates scale-invariant.
Therefore, for graphical representation purposes,
the correlation-based version will be preferred.

Plotting on a display usually requires providing
coordinates in a rectangular Cartesian system.
Therefore, we will need to compute the equivalent rectangular coordinates
of our general, possibly non-rectangular coordinates.
This implies pre-multiplying its coordinates in basis $`r TEST_SPACE_BASIS`$
by $`r TRANSFORM_MATRIX_TRANSP_INV`$, to find
the equivalent, orthonormalized coordinates.
Therefore, we will need to define
a convenient value for $`r TRANSFORM_MATRIX_TRANSP_INV`$
that allows us to represent the test space structure and the items
in a rectangular Cartesian system.

@@

As plotting more than two dimensions is difficult on a bidimensional display
(and hardly possible at all for more than three dimensions with the
currently available technology),
we will limit ourselves to the bidimensional case here.

The example in Figure \@ref(fig:item-plot-out) showcases a representation of
a set of items in two different bidimensional spaces.
In the first case, the two latent dimensions are independent,
with $`r TRANSFORM_MATRIX_TRANSP_INV`$ being an identity matrix.
In the second one, the correlation between the dimensions is $`r CORR_OBL_OUT`$.
To keep the horizontal axis invariant,
we define $`r TRANSFORM_MATRIX_TRANSP_INV`$ such that
it only rotates the vertical axis [@harman_modern_1970]:

@@

@reckase_difficulty_1985 proposed a definition of the $`r MIL_PARAM`$ parameter
general enough to "be used with any model that yields probabilities that
increase monotonically with an increase in ability on any dimension" (p. 411).
However, we have discussed that this definition implicitly assumes
the orthonormality of the latent space.
Applying the general procedure outlined by @reckase_difficulty_1985 and later
extended by @reckase_discriminating_1991
we have obtained a set of multidimensional parameters
that generalize the original results in two aspects:
(1) to a non-orthonormal space that accounts for the covariance structure of
the multivariate latent variable, and (2) to any case of unchanging monotony.
The latter may seem superfluous, given that several examples in the literature
already make use of this generalization
[e.g., the one we have used to illustrate it: @tezza_modelo_2018].
However, this generalization had never been formalized, up to now.
Therefore, we highlight here the interest of showing how
the item psychometric properties are paralleled by their geometrical properties,
and how this holds for any case of unchanging monotony
of the item response function.
Regarding the first generalization,
it is worth noting that @zhang_theoretical_1999 already proposed a formulation
of the $`r MDISC_SYM`$ equivalent to our covariance-based version,
but for a more general, semi-parametric formulation
(although constrained to non-decreasing monotony).
However, they provided no formal proof for this result either.
Our results provide this formal derivation, although only for the specific
case of the `r MODEL_ACRONYM` model.

As we have seen, taking the latent space structure into account may have a
substantial effect in the resulting values and interpretation of the parameters,
something not contemplated by the original
(hereby referred to as agnostic) version.
Moreover, in applications where the latent space structure is critical
(e.g., multi-group IRT, equating, and linking),
the agnostic version may overlook relevant differences in
the values of the multidimensional parameters.
Thus, the covariance structure of the latent space should be considered
when computing the multidimensional parameters of a MIRT model.

With that purpose in mind, we have defined two new versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based parameter has the drawback of not fulfilling
the first of @reckase_discriminating_1991's properties, namely,
the scale invariance property.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
More importantly, the covariance-based version
of the multidimensional parameters is invariant to
changes of latent-space basis.
Once again, this is important in applications where
potentially different bases are involved.
Given these considerations,
we propose to adopt the covariance-based version of
the multidimensional parameters as the most general definition,
and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of being scale-invariant,
which is convenient for fidelity when representing the items.
However, it is worth noting that it has been defined ad-hoc
to fulfill this specific property.
Therefore, it lacks other desirable properties of the covariance-based version,
like being invariant to a latent space scaling transformation,
or yielding a diagonal latent covariance matrix in the corresponding
orthogonal latent space basis.
Therefore, we recommend its use for graphical representation purposes only.
As for the agnostic version, we recommend it only in
the (unlikely) case of having no information about the latent space structure,
or when such structure is considered irrelevant
(e.g., plotting the discrimination parameters after an oblique rotation).

One may consider that the formal derivations presented here are
overly complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
@reckase_difficulty_1985's and @reckase_discriminating_1991's procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
Such a procedure will of course yield identical results,
given the invariance of the parameters to a change of basis.
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.
Computing the parameters in the orthonormalized (or just orthogonalized) space,
by contrast, would require computing the cosines in
the original basis afterwards.

On the other hand, establishing the relationship between the original and
the orthonormalized basis has provided us with an important insight:
The distinction between the latent space and the test space.
Our derivations show that, to properly represent the items,
we need to define a vector space that has
the latent covariance (or correlation) matrix as its inner product matrix.
On the other hand, the latent space,
where the person parameters are represented,
has the inverse of the covariance (correlation) matrix as
its inner product matrix.
This implies that the axes defined by the test space basis,
and not the latent space basis,
are a geometrical representation of the covariance structure.
This result is both surprising and counterintuitive
---Several instances in the MIRT literature suggest that
"the angle between the axes [of the latent ability space] represents
the degree of correlation" [@ackerman_multidimensional_2005-1, p. 16].
We hope our results help shed a clearer light on
this often misunderstood topic.

It is worth noting that a similar relationship exists in
the common factor literature
between the *primary factors* and Thurstone's *reference axes*
[@harman_modern_1970, see p. 291].
Nevertheless, one must not make the mistake of assimilating the test space and
the *common-factor space*.
Their parameterization, albeit related [@mcdonald_test_1999], is different,
and we do not mean to imply that the derivations made here are directly
generalizable to the common factor model.
Those relationships deserve further exploration on their own.

Finally, we must also warn against generalizing these results
to other IRT models without further consideration.
First, we have only considered the case of dichotomous items,
whereas non-cognitive applications often (and sometimes cognitive ones as well)
require modeling several response categories.
Although proposals such as the graded-scale [@samejima_estimation_1968] and
the nominal response [@bock_estimating_1972] model are closely related to
the 2-parameter logistic model, it would be hasty to assume that these
parameters generalize in any way to
the multidimensional versions of those models without a formal proof.
The same can be said from the multidimensional parameters of other models
for dichotomous items different from the `r MODEL_ACRONYM` model.
We expect this work to help set the foundations for investigating those
generalizations.
Hopefully, this will contribute to a better understanding
of the underlying multidimensional measurement theory, its implications,
and its interpretation.

>>>>

We must make a few remarks here. First, as we are not constraining ourselves to
cognitive traits, we speak of _positive_ instead of _correct response_.
This may imply a correct response, if we are talking about a cognitive test, or
endorsing an item (or a certain response option) in a non-cognitive domain.
Second, although not explicitly stated, $`r DISCR_VECTOR`$ is a vector of
non-negative real numbers in Reckase
[-@reckase_difficulty_1985, a decrease in the probability of a correct response
with an increase in ability makes little sense].
In our case, $`r DISCR_VECTOR`$ is a vector of real numbers; the value
of the $`r DIM_INDEX`$-th component $`r DISCR_PARAM`$ can be negative if
the item is *inverse* for (i.e., taps the negative pole of)
latent dimension $`r DIM_INDEX`$.
Finally, as responses are not necessarily interpreted as correct or incorrect,
parameter $`r INTERCEPT_PARAM`$ is said to be related to the item _location_,
instead of the item difficulty.

Our aim is to find the point of maximum slope of $`r IRF_ABBR`$ in
$`r LATENT_SPACE`$.
Because the slope is dependent on the direction considered,
we need to reparameterize Equation \@ref(eq:M2PL-formulation)
in polar coordinates, as in @reckase_difficulty_1985.

However, Reckase's Equation 3 implicitly assumes orthogonality.
In order to contemplate the general case,
we need to introduce a few linear algebra results before proceeding further.

@@

Thus, inner product $`r LS_BASIS_INNER_PROD`$ has $`r INNER_PROD_MAT_EQ`$ as
Gram matrix, i.e., $`r INNER_PROD_EQ`$, with $`r INNER_PROD_MAT_ELEMENT_EQ`$ as
the $`r INNER_PROD_MATRIX_INDEX`$-th element of $`r INNER_PROD_MATRIX`$. As
$`r TRANSFORM_MATRIX`$ (and consequently $`r TRANSFORM_MATRIX_TRANSP`$) is
invertible, so $`r INNER_PROD_MATRIX`$ is, and $`r INNER_PROD_MATRIX_INV_EQ`$.

@@

1. $`r TRAIT_VEC_ORTH_POLAR_EQ`$

1. $`r TRAIT_VECTOR_POLAR_EQ`$

1. $`r TRAIT_VEC_ALT_POLAR_EQ`$

Given (1),

@@

Using (2) and (3) we get that

@@

Applying Proposition 1,

@@

with $`r DIAG_MATRIX_INNER_PROD_INV`$ a diagonal matrix where
$`r MATRIX_INNER_PROD_INV_EL_DIAG_EQ`$.

@@

The second property derives straightforwardly from
Equation \@ref(eq:distance-param),
and the third one is implicit in the definition of the $`r MDISC_SYM`$.
Simple arithmetic can show that the first holds true whenever the
$`r DIAG_KTH_ELEMENT`$-th element of $`r INNER_PROD_MATRIX_INV`$ equals 1.
Therefore, both the agnostic and the correlation-based version fulfill it.
$`r MDISC_COV_PARAM`$ does not though,
as $`r COV_ELEMENT`$ is not generally equal to 1.
Nevertheless, it can be interpreted as a *scaled version* of the index,
with the standard deviation of the corresponding dimension as scaling factor.
This transformation may be made in the unidimensional case as well,
thus finding a *standard-metric* discrimination parameter
(i.e., referred to a unitary variance space).

The (signed) distance to the origin,
given by the second element of $`r MIL_PARAM`$, has an interesting property:
By the definition of the inner product,
$`r MDISC_SYM`$ is strictly positive for non-null vectors.
Therefore, the sign of $`r DISTANCE_PARAM`$ is equal to
the sign of $`r INTERCEPT_PARAM`$.
The implication is that the item is displaced along
its measurement direction forwards or backwards with respect to the origin.

The measurement direction is given by the signs of $`r DIR_COS_ITEM_VEC`$ which,
also due to the strict positivity of $`r MDISC_SYM`$,
are equal to the signs of $`r DISCR_VECTOR`$.
This property leads directly to generalizing the indices to the monotonically
non-increasing case: The measurement direction relative to dimension
$`r DIM_INDEX`$ will be negative when $`r IRF_ABBR`$ is
monotonically decreasing with respect to variations along $`r TRAIT_COMPONENT`$.
When $`r IRF_ABBR`$ is constant with respect to
variations along $`r TRAIT_COMPONENT`$, $`r SIGN_COS_VEC_ITEM_EQ`$,
which means that $`r MIL_PARAM`$ is orthogonal to the $`r DIM_INDEX`$-th axis.
However, note that this does not necessarily imply that the vector is
parallel to any other axis or strictly contained in the hyperplane formed
by other axes; this will always be true for the $`r MIL_AG_PARAM`$ due to the
orthogonality assumption, but it will depend on the correlation matrix
$`r CORR_MATRIX`$ for the other two versions of the index.

@@

The aforementioned properties are easier to apprehend with
a visual representation of the items.
A vector plot [@ackerman_creating_1994] is the most appropriate way of
representing the items graphically;
it allows one to plot several items altogether and analyze them
in terms of their multidimensional parameters [@ackerman_graphical_1996]:
An item vector will be applied at a direction and (signed) distance
from the origin of the coordinate system given by its $`r MIL_PARAM`$ parameter
and will have a length equal to its $`r MDISC_SYM`$ parameter.

As plotting more than two dimensions is difficult on a bidimensional display
(and hardly possible at all for more than three dimensions with the
currently available technology),
we will limit ourselves to the bidimensional case here.
Plotting on a two-dimensional display usually requires providing
coordinates in a rectangular Cartesian system.
Therefore, we will need to compute the equivalent rectangular coordinates
of our general, possibly non-rectangular coordinates.
However, this implies pre-multiplying by $`r TRANSFORM_MATRIX`$,
instead of $`r TRANSFORM_MATRIX_TRANSP_INV`$ as would be expected from
Equation \@ref(eq:discr-vector-std).
This is because computing the equivalent rectangular coordinates
of the oblique ones is different from computing the item parameters
transformed to an orthogonal space.
While the latter implies rotating the axes to an orthogonal configuration and
computing the new coordinates in these axes,
the former implies transforming the space altogether along with the
axes transformation.

Therefore, we need to define a convenient value for $`r TRANSFORM_MATRIX`$
that allows us to represent the test space structure
and the items by transforming their coordinates to
a rectangular Cartesian system.
However, to facilitate the interpretation,
we want to avoid changing the scaling of the parameters.
We will thus prefer the correlation-based over the covariance-based indices.
Thus, we define $`r TRANSFORM_MATRIX`$ such that
$`r TRANSFORM_MATRIX_TRANSP_INV`$ is a square root matrix of $`r CORR_MATRIX`$.

@@

The example in Figure \@ref(fig:item-plot-out) showcases a representation of
a set of items in two different bidimensional spaces.
In the first case, the two latent dimensions are independent,
with $`r TRANSFORM_MATRIX`$ being an identity matrix.
In the second one, the correlation between the dimensions is $`r CORR_OBL_OUT`$.
In order to keep the horizontal axis invariant,
we define $`r TRANSFORM_MATRIX`$ such that it only rotates the vertical axis
[@harman_modern_1970]:

@@

@reckase_difficulty_1985 proposed a definition of the $`r MIL_PARAM`$ parameter
general enough to "be used with any model that yields probabilities that
increase monotonically with an increase in ability on any dimension" (p. 411).
However, we have discussed that this definition implicitly assumes
the orthonormality of the latent space.
Applying the general procedure outlined by @reckase_difficulty_1985 and later
extended by @reckase_discriminating_1991
we have obtained a set of multidimensional parameters
that generalize the original results in two aspects:
(1) to a non-orthonormal space that accounts for the covariance structure of
the multivariate latent variable, and (2) to any case of unchanging monotony.
It is worth noting that @zhang_theoretical_1999 already proposed a formulation
of the $`r MDISC_SYM`$ equivalent to our covariance-based version,
but for a more general, semi-parametric formulation
(although constrained to non-decreasing monotony).
However, they provided no formal proof for this result.
Our results provide this formal derivation, although only for the specific
case of the `r MODEL_ACRONYM` model.

As we have seen, taking the latent space structure into account may have a
substantial effect in the resulting values and interpretation of the parameters,
something not contemplated by the original
(hereby referred to as agnostic) version.
Moreover, in applications where the latent space structure is critical
(e.g. multi-group IRT, equating, and linking),
the agnostic version may overlook relevant differences in
the values of the multidimensional parameters.
Thus, the covariance structure of the latent space should be considered
when computing the multidimensional parameters of a MIRT model.

With that purpose in mind, we have defined two versions of these parameters:
One that takes into account the whole covariance structure,
and another one that only considers the correlations among latent dimensions.
The covariance-based index has the drawback of not fulfilling
the first of @reckase_discriminating_1991's properties, namely,
that $`r MDISC_UNIDIM_EQ`$ if an item only measures dimension $`r DIM_INDEX`$.
However, we have provided a rationale for the violation of this property,
arguing that it is also paralleled in unidimensional IRT.
Given its equivalence to the inner product definition in a space with
an identity covariance matrix,
it also refers the item parameters to an orthonormal metric,
comparable among different latent spaces.
Once again, this is important in applications where
potentially different latent spaces are involved.
Given these considerations, we propose to adopt the covariance-based version of
the multidimensional parameters and refer to them simply as
the $`r MDISC_SYM`$ and $`r MIL_PARAM`$ parameters.
The correlation-based version, on the other hand,
has the advantage of preserving the metric of the parameters,
which is convenient for fidelity when representing the items.
Therefore, we recommend using the correlation-based version of the parameters
for graphical representation purposes only.

One may consider that the formal derivations presented here are overly
complicated,
when one might simply transform the space to rectangular Cartesian coordinates
with Equations \@ref(eq:change-basis) and \@ref(eq:discr-vector-std) and
then compute the parameters in the transformed space using
@reckase_difficulty_1985's and @reckase_discriminating_1991's procedure.
After all, interpreting the axes of this space as substantive latent traits
depends on "the distinction between coordinate axes and the constructs that are
the target of the instrument"
(Reckase, M.D., personal communication, 28 April 2015).
However, the computations in the oblique space allow us to obtain
the direction cosines relative to the original latent dimensions,
which may give relevant information if these are actual, substantive traits.
Moreover, strictly following @reckase_difficulty_1985's method in the original,
oblique space has also provided us with an important insight:
The distinction between the latent space and the test space.
Our derivations show that, in order to properly represent the items,
we need to define a vector space that has
the latent covariance (or correlation) matrix as its inner product matrix.
On the other hand, the latent space,
where the person parameters are represented,
has the inverse of the covariance (correlation) matrix as
its inner product matrix.
It is worth noting that a similar relationship exists in
the common factor literature
between the *primary factors* and Thurstone's *reference axes*
[@harman_modern_1970, see p. 291].
Nevertheless, one must not make the mistake of assimilating the test space and
the *common-factor space*.
Their parameterization, albeit related [@mcdonald_test_1999], is different,
and we do not mean to imply that the derivations made here are directly
generalizable to the common factor model.
Those relationships deserve further exploration on their own.

Finally, we must also warn against generalizing these results
to other IRT models without further consideration.
First, we have only considered the case of dichotomous items,
whereas non-cognitive applications often (and sometimes cognitive ones as well)
require modeling several response categories.
Although models such as the graded-scale [@samejima_estimation_1968] and
the nominal response [@bock_estimating_1972] model are closely related to
the 2-parameter logistic model, it would be hasty to assume that these
parameters generalize in any way to
the multidimensional versions of those models without a formal proof.
The same can be said from the multidimensional parameters of other models
for dichotomous items different from the `r MODEL_ACRONYM` model.
We expect this work to help set the foundations for investigating those
generalizations.
Hopefully, this will contribute to a better understanding
of the underlying measurement theory, its implications and its interpretation.
